---
title: "Data Manipulation"
author: "ruby"
date: "2025-09-18"
output: github_document
---

```{r}
library(tidyverse)
```


let's import the datasets. 

```{r}
litters_df = 
  read_csv("data/FAS_litters.csv", na = c("NA", ".", ""))

litters_df = 
  janitor::clean_names(litters_df)

pups_df =
    read.csv("data/FAS_pups.csv", skip = 3, na = c("NA", ".", ""))

pups_df = 
  janitor::clean_names(pups_df)
```

## `select`

`select` selects variables from my dataframe 

```{r}
select(litters_df, group, litter_number, gd0_weight)
```

select a range

```{r}
select(litters_df, group:gd_of_birth)
```

select by removal 

```{r}
select(litters_df, -group)
select(litters_df, -(group:gd_of_birth))
```

select by prefix

```{r}
select(litters_df, group, starts_with("gd"))
```

rearrange by selecting -> however you write it is how select will present it to you

```{r}
select(litters_df, starts_with("gd"), group)
select(litters_df, litter_number, everything())

relocate(litters_df, litter_number)
```

relocate does the same thing as the select 

rename by selecting 

```{r}
select(litters_df, GROUP = group, everything())
rename(litters_df, GROUP = group)
```

what if i only select on variable 
select gives you a dataframe, pull gives you a vector

```{r}
select(litters_df, group)

pull(litters_df,group)
```

learning assessment: in the pups dataset, select the columns containing litter number, sex, and pd ears 

```{r}
select(pups_df, litter_number, sex, pd_ears)
```

## `filter`

use `filter()` to filter out rows. 

```{r}
filter(litters_df, gd_of_birth == 20, pups_born_alive == 6)
```

look for more than two pups alive

```{r}
filter(litters_df, pups_born_alive > 5)
filter(litters_df, pups_born_alive >= 5)
filter(litters_df, pups_born_alive <= 5)
```

look for non-matches 

```{r}
filter(litters_df, pups_born_alive != 6)
```

match to a set

```{r}
filter(litters_df, group %in% c("Con7", "Con8"))
```

filter out missing rows with `drop_na()`

```{r}
drop_na(litters_df)
drop_na(litters_df, gd0_weight)
```

learning assessment: 

```{r}
filter(pups_df, sex == 1)
filter(pups_df, pd_walk < 11, sex == 2)
```

## `mutate`

use `mutate()` to create or modify variables. 

```{r}
mutate(litters_df, 
       wt_gain = gd18_weight - gd0_weight,
       group = str_to_lower(group)
       )
```

## `arrange`

`arrange()` arranges things

```{r}
arrange(litters_df, desc(pups_born_alive))
arrange(litters_df, desc(group), pups_born_alive)

arrange(litters_df, gd0_weight)
```


## PIPES 

let's see a bad way to do things (very clunky, & messy)

```{r}
litters_df = 
  read_csv("data/FAS_litters.csv", na = c("NA", ".",""))

litters_df = 
  janitor::clean_names(litters_df)

litters_df = 
  select(litters_df, group, litter_number, starts_with("gd"))

litters_df = 
  drop_na(litters_df)

litters_df = 
  mutate(
    litters_df,
    wt_gain = gd18_weight - gd0_weight
  )
```

Use pipes to avoid all this!! (cmd-shift-m for pipe operator)

```{r}
litters_df = 
  read_csv("data/FAS_litters.csv", na = c("NA", ".", "")) |> 
  janitor::clean_names() |> 
  select(group, litter_number, starts_with("gd")) |> 
  drop_na() |> 
  mutate(
    wt_gain = gd18_weight - gd0_weight
  )
```

learning assessment: import and tidy pups data

```{r}
pups_df = 
  read_csv("data/FAS_pups.csv", skip = 3, na = c("NA", ".", "")) |> 
  janitor::clean_names() |> 
  filter(sex == 1) |> 
  select(-pd_ears) |> 
  mutate(
    pd_pivot_ge7 = pd_pivot >= 7
  )
```

