---
title: "Data Manipulation"
author: "ruby"
date: "2025-09-18"
output: github_document
---

```{r}
library(tidyverse)
```


let's import the datasets. 

```{r}
litters_df = 
  read_csv("data/FAS_litters.csv", na = c("NA", ".", ""))

litters_df = 
  janitor::clean_names(litters_df)

pups_df =
    read.csv("data/FAS_pups.csv", skip = 3, na = c("NA", ".", ""))

pups_df = 
  janitor::clean_names(pups_df)
```

## `select`

`select` selects variables from my dataframe 

```{r}
select(litters_df, group, litter_number, gd0_weight)
```

select a range
```{r}
select(litters_df, group:gd_of_birth)
```

select by removal 
```{r}
select(litters_df, -group)
select(litters_df, -(group:gd_of_birth))
```

select by prefix
```{r}
select(litters_df, group, starts_with("gd"))
```

rearrange by selecting -> however you write it is how select will present it to you

```{r}
select(litters_df, starts_with("gd"), group)
select(litters_df, litter_number, everything())

relocate(litters_df, litter_number)
```

relocate does the same thing as the select 

rename by selecting 
```{r}
select(litters_df, GROUP = group, everything())
rename(litters_df, GROUP = group)
```

what if i only select on variable 
select gives you a dataframe, pull gives you a vector
```{r}
select(litters_df, group)

pull(litters_df,group)
```

learning assessment: in the pups dataset, select the columns containing litter number, sex, and pd ears 
```{r}
select(pups_df, litter_number, sex, pd_ears)
```

## `filter`
