---
title: "Data Import"
output: html_document
---

```{r}
library(tidyverse)
library(readxl)
library(haven)
```

let's import a dataset. 

```{r}
litters_df =
  read_csv("data/FAS_litters.csv")
```

Look at your data!

```{r}
names(litters_df)
```

Update the names in `litters_df`.

```{r}
litters_df = 
  janitor::clean_names(litters_df)

names(litters_df)
```

Look at the data

```{r}
litters_df
```

Sometimes skimming data is net

```{r}
skimr::skim(litters_df)
```


##fix the missingness

```{r}
litters_df = 
  read_csv("data/FAS_litters.csv", na = c("NA", ".",""), skip = 5)
```


##import pups data 

```{r}
pups_df = 
  read_csv(
    "data/FAS_pups.csv", 
    skip = 3,
    na = c("NA", ".", "")
    )

pups_df = 
  janitor::clean_names(pups_df)
```

## what about excel files 

csvs are great, but sometimes you get an excel file. 

```{r}
mlb_df =
  read_excel("data/mlb11.xlsx")
```

Import LotR word counts. 

```{r}
fotr_df = 
  read_excel("data/LotR_Words.xlsx", range = "B3:D6")
```


## SAS!

Import the PULSE data

```{r}
pulse_df = 
  read_sas("data/public_pulse_data.sas7bdat")

pulse_df = 
  janitor::clean_names(pulse_df) 
```

## Why is read.csv not good 
read.csv is base R functions - no tidyverse functions
```{r, eval = FALSE}
litters_df_base = 
  read.csv("data/FAS_litters.csv")
```

## Data exporting

```{r}
write_csv(fotr_df, "data/fotr_df.csv")
```

